---
title: "Getting started with Ritable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with Ritable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(Ritable)
```

## Overview

**Ritable** estimates narrow-sense heritability (h²) for quantitative traits in
family cohort studies using a profile-likelihood variance-components approach,
without requiring SOLAR Eclipse or any proprietary software.

The three main steps are:

1. Build the additive genetic relationship matrix (GRM) from a pedigree.
2. Estimate heritability for one trait with `herit_vc()`.
3. Scale to many traits with `herit_batch()`, then visualise with `plot_forest()`.

---

## 1. Build the GRM

Your pedigree data frame needs four columns: individual ID, father ID,
mother ID, and sex (1 = male, 2 = female). Missing parents are coded `NA` or `0`.

```{r build-grm}
# Minimal example pedigree (founders + one generation of offspring)
ped <- data.frame(
  id  = 1:8,
  pat = c(0, 0, 0, 0, 1, 1, 3, 3),
  mom = c(0, 0, 0, 0, 2, 2, 4, 4),
  sex = c(1, 2, 1, 2, 1, 2, 1, 2)
)

# Study subjects are the offspring (IDs 5-8)
A <- build_grm(ped, study_ids = 5:8)
round(A, 3)
```

The diagonal is 1 for non-inbred individuals. Full siblings share 0.5 of their
genetic material on average, so off-diagonal entries for sibling pairs are 0.5.

---

## 2. Estimate heritability for a single trait

```{r single-trait}
set.seed(42)

# Simulate some phenotype data for the study subjects
dat <- data.frame(
  IID     = 5:8,
  age     = c(35, 38, 40, 44),
  sex_num = c(1, 2, 1, 2),
  bmi     = c(24.1, 27.3, 22.8, 29.5)
)

# Unadjusted model
res_unadj <- herit_vc("bmi", grm = A, data = dat, min_n = 3)

# Adjusted model (age + sex)
res_adj <- herit_vc("bmi", grm = A, data = dat,
                    covs  = c("age", "sex_num"),
                    label = "bmi_adj",
                    min_n = 3)

str(res_unadj)
```

The returned list contains:

| Field | Description |
|-------|-------------|
| `h2` | MLE narrow-sense heritability |
| `se` | SE from profile-LL curvature |
| `ci_lo`, `ci_hi` | 95% profile-likelihood CI |
| `pval` | One-sided LRT p-value (boundary-corrected) |
| `sigma2_a` | Additive genetic variance |
| `sigma2_e` | Residual variance |

---

## 3. Batch estimation

For real analyses you will typically run many traits across several covariate
models. `herit_batch()` handles this and returns a tidy data frame.

```{r batch}
# Add a second trait
dat$hdl <- c(55.0, 60.2, 48.3, 72.1)

covs_list <- list(
  unadj = NULL,
  cov1  = c("age", "sex_num")
)

res <- herit_batch(
  traits    = c("bmi", "hdl"),
  grm       = A,
  data      = dat,
  covs_list = covs_list,
  min_n     = 3,
  .progress = FALSE
)

res
```

---

## 4. Forest plot

```{r forest, fig.width=7, fig.height=3}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  plot_forest(res,
              model_filter  = "cov1",
              colour_by     = "trait",
              title         = "Adjusted heritability (age + sex)")
}
```

---

## 5. Real-study workflow

A typical family cohort workflow looks like this:

```{r real-workflow, eval=FALSE}
library(Ritable)

# 1. Load pedigree (full pedigree including founders)
ped <- read.csv("baependi_pedigree_full.csv")

# 2. Load analysis dataset
dat <- read.csv("baependi_analysis_dataset.csv")
dat$age2    <- dat$age^2
dat$age_sex <- dat$age * dat$sex_num

# 3. Build GRM once -- reuse for all traits
A <- build_grm(ped,
               study_ids = dat$IID,
               id_col    = "IID",
               pat_col   = "pat",
               mom_col   = "mat",
               sex_col   = "sex_num")

# 4. Define covariate models
covs_list <- list(
  unadj = NULL,
  cov1  = c("age", "sex_num"),
  cov2  = c("age", "sex_num", "age2"),
  cov3  = c("age", "sex_num", "age2", "age_sex")
)

# 5. Run batch estimation
brain_traits <- c("k", "S", "I", "AvgThickness", "localGI_corrected")

res <- herit_batch(
  traits    = brain_traits,
  grm       = A,
  data      = dat,
  covs_list = covs_list
)

# 6. Save and plot
write.csv(res, "table_heritability_brain.csv", row.names = FALSE)
plot_forest(res, model_filter = "cov2",
            title = "Brain morphology heritability (age + sex + age2)")
```

---

## Citation

If you use **Ritable** in a publication, please cite:

```
Franca, L. & Leocadio-Miguel, M. (2026). Ritable: Pedigree-Based Heritability
Estimation for Family Cohort Studies. R package version 0.1.0.
https://github.com/circadia-bio/R-itable
```
