---
title: "Twin cohorts: MZ relatedness, household effects, and bivariate correlations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Twin cohorts: MZ relatedness, household effects, and bivariate correlations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(Ritable)
```

## Overview

The [Getting started](getting-started.html) vignette covers `herit_vc()` and
`herit_batch()`: single-component (additive genetic, A-only) heritability for
general family pedigrees. Twin studies add three things that vignette doesn't
cover:

1. **Monozygotic (MZ) twins are genetically identical** (relatedness 1.0, not
   the ordinary full-sibling 0.5) -- `build_grm()`'s `mz_col` argument.
2. **Twins reared together share more than genes** -- a common/shared
   environment ("household", or **C** in the classic **ACE** model) that can
   be confounded with additive genetic (**A**) effects unless you explicitly
   test for it -- `build_household()` and `herit_ace()`.
3. **Twin designs are the classical setting for asking whether two traits
   share genetic or environmental causes** -- bivariate genetic/environmental
   correlation, via `herit_bivar()` and `herit_bivar_batch()`.

This vignette builds a small synthetic twin cohort and walks through all
three.

---

## 1. Simulate a twin cohort

A twin pedigree looks like any other pedigree, but with two extra columns
`build_grm()` and `build_household()` care about: an **MZ-twin label**
(shared only by monozygotic pairs) and a **household ID** (shared by anyone
reared together, MZ or DZ).

We'll simulate 30 MZ pairs and 30 DZ pairs (120 individuals total), each
pair with its own synthetic founder "parents" (`build_grm()` needs founders
present as their own rows, even when -- as here -- they aren't real study
participants). Two correlated phenotypes are simulated with known genetic and
environmental correlations, so we have a known answer to check `herit_bivar()`
against later.

```{r simulate}
# Tiny base-R replacement for MASS::mvrnorm for a 2x2 covariance -- R-itable
# has no dependency on MASS, so the vignette doesn't introduce one either.
rbinorm <- function(n, rho) {
  z <- matrix(rnorm(2 * n), n, 2)
  L <- chol(matrix(c(1, rho, rho, 1), 2, 2))
  z %*% L
}

set.seed(2026)
n_mz <- 30
n_dz <- 30
n_pairs <- n_mz + n_dz

rhoG_true <- 0.85  # true genetic correlation (before dilution from DZ noise)
rhoE_true <- 0.35  # true environmental correlation

ped_rows   <- vector("list", n_pairs)
pheno_rows <- vector("list", n_pairs)

# Family-level genetic latents for the two traits, correlated at rhoG_true
fam_g <- rbinorm(n_pairs, rhoG_true)

for (i in seq_len(n_pairs)) {
  is_mz  <- i <= n_mz
  fa     <- paste0("F", i)
  mo     <- paste0("M", i)
  id1    <- (i - 1L) * 2L + 1L
  id2    <- id1 + 1L
  mztwin <- if (is_mz) paste0("MZ", i) else NA_character_
  hh     <- paste0("HH", i)
  pair_sex <- if (is_mz) rep(sample(1:2, 1), 2) else sample(1:2, 2, replace = TRUE)

  ped_rows[[i]] <- data.frame(
    id = c(id1, id2), pat = c(fa, fa), mom = c(mo, mo),
    mztwin = c(mztwin, mztwin), zygosity = if (is_mz) "MZ" else "DZ",
    sex = pair_sex, hhid = c(hh, hh), stringsAsFactors = FALSE
  )

  # Additive genetic effect: shared fully within MZ pairs; shared "family"
  # component plus individual-specific noise within DZ pairs (average
  # relatedness 0.5).
  g1_fam <- fam_g[i, 1]; g2_fam <- fam_g[i, 2]
  if (is_mz) {
    g1 <- c(g1_fam, g1_fam)
    g2 <- c(g2_fam, g2_fam)
  } else {
    g1 <- g1_fam + rnorm(2)
    g2 <- g2_fam + rnorm(2)
  }

  # Environmental effect, correlated at rhoE_true, independent across twins
  env <- rbinorm(2, rhoE_true)

  y1 <- sqrt(0.5) * g1 + sqrt(0.5) * env[, 1]
  y2 <- sqrt(0.4) * g2 + sqrt(0.6) * env[, 2]

  pheno_rows[[i]] <- data.frame(id = c(id1, id2), sleep_quality = y1, anxiety = y2)
}

ped_df   <- do.call(rbind, ped_rows)
pheno_df <- do.call(rbind, pheno_rows)

# build_grm() needs founders present as their own rows (pat/mom = NA)
fathers  <- unique(ped_df$pat)
mothers  <- unique(ped_df$mom)
founders <- data.frame(
  id = c(fathers, mothers), pat = NA, mom = NA, mztwin = NA, zygosity = NA,
  sex = c(rep(1, length(fathers)), rep(2, length(mothers))), hhid = NA,
  stringsAsFactors = FALSE
)
full_ped <- rbind(founders, ped_df)

head(ped_df)
```

---

## 2. Build the GRM with MZ relatedness

Without `mz_col`, `build_grm()` treats every pair as ordinary full siblings
(relatedness 0.5) -- correct for DZ twins, wrong for MZ twins. Pass `mz_col`
to fix that:

```{r grm}
A <- build_grm(full_ped, study_ids = ped_df$id, mz_col = "mztwin")

mz_ids <- ped_df$id[ped_df$zygosity == "MZ"][1:2]
dz_ids <- ped_df$id[ped_df$zygosity == "DZ"][1:2]

A[as.character(mz_ids), as.character(mz_ids)]
A[as.character(dz_ids), as.character(dz_ids)]
```

The MZ pair's off-diagonal is `1.0`; the DZ pair's is the ordinary
full-sibling `0.5`. `mz_col` also handles larger groups (e.g. MZ triplets) --
every pairwise combination within a labelled group is marked, not just the
first two.

`build_household()` builds the companion matrix for the shared-environment
("household") effect, from the `hhid` grouping column:

```{r household}
C <- build_household(ped_df$hhid, ped_df$id)
C[as.character(mz_ids), as.character(mz_ids)]
```

---

## 3. A vs C identifiability: `herit_ace()`

The classical twin design lets you decompose phenotypic variance into
**A**dditive genetic, **C**ommon/shared environment, and unique
**E**nvironment components -- but only if the data actually support telling A
and C apart. `herit_ace()` fits all four nested models (sporadic, AE-only,
CE-only, full ACE) and gives you the two comparisons that matter:

```{r ace}
dat <- pheno_df
dat$IID <- dat$id

res_ace <- herit_ace("sleep_quality", grm = A, household = C, data = dat, id_col = "IID")
str(res_ace)
```

Reading the output:

- **`p_c_vs_sporadic`** -- is there a familial (shared) effect at all,
  ignoring for a moment whether it's genetic or environmental? A small
  p-value here means yes.
- **`p_ace_vs_ae`** -- once additive genetic effects are in the model, does
  adding a separate household term improve the fit? If not (large p-value,
  and/or `c2_ace` at or near `0`), A and C are **not separately identifiable**
  in this sample -- the household signal detected above gets fully absorbed
  into the genetic term when both are estimated jointly. This is a common,
  real finding in twin studies with modest samples, not a bug: report the
  simpler AE model in that case (`h2_ae`), and note the limitation.

### Covariates and automatic screening

`herit_ace()` can adjust for covariates via `covs`, with optional automatic
screening that replicates SOLAR Eclipse's `polygenic -screen` algorithm
(confirmed from SOLAR's own source, not assumed): fit the AE model with
*all* candidate covariates, then likelihood-ratio-test each one
individually against that full model, and drop any with p >= `prob_level`
(default `0.10`, SOLAR's own default). The same final covariate set is
then used for all four nested models, keeping the sporadic/AE/CE/ACE
comparisons validly nested.

```{r ace-covs}
# A synthetic covariate that genuinely predicts sleep_quality, and one
# that's pure noise -- screening should keep the first and drop the second
# (though with any single random draw, a pure-noise covariate can
# occasionally clear a p < 0.10 threshold by chance, same as any
# significance test).
set.seed(1)
dat$age       <- round(runif(nrow(dat), 20, 60))
dat$predictor <- dat$sleep_quality * 2 + rnorm(nrow(dat), sd = 0.1)

res_ace_cov <- herit_ace("sleep_quality", grm = A, household = C, data = dat,
                         covs = c("age", "predictor"), id_col = "IID")
res_ace_cov$covariate_lrt_p  # each candidate's individual LRT p-value
res_ace_cov$covs_kept        # the ones that survived screening
```

Pass `screen = FALSE` to include every covariate in `covs` unconditionally,
skipping the LRT screening step entirely.

---

## 4. Bivariate genetic/environmental correlation: `herit_bivar()`

`herit_bivar()` asks a different question: for **two** traits, how much of
their phenotypic correlation is explained by shared genetic causes (rhoG) vs
shared environmental causes (rhoE)?

```{r bivar}
res_biv <- herit_bivar("sleep_quality", "anxiety", grm = A, data = dat, id_col = "IID")
str(res_biv)
```

We simulated `rhoG_true = 0.85` (diluted somewhat by the individual-specific
genetic noise added for DZ pairs) and `rhoE_true = 0.35`; `herit_bivar()`
recovers a clearly significant genetic correlation and a smaller,
non-significant environmental one -- a realistic pattern (a trait pair that
looks correlated mainly for genetic, not environmental, reasons).

`rhoP` is the derived phenotypic correlation implied by rhoG, rhoE, and the
two univariate heritabilities:

```{r rhop-check}
with(res_biv, rhoG * sqrt(h2_1 * h2_2) + rhoE * sqrt((1 - h2_1) * (1 - h2_2)))
res_biv$rhoP
```

---

## 5. Many trait pairs at once: `herit_bivar_batch()`

Real analyses usually involve one trait of interest against a panel of
others. `herit_bivar_batch()` iterates `herit_bivar()` over a set of pairs and
applies Benjamini-Hochberg FDR correction separately within each of the three
p-value families (rhoG, rhoE, rhoP) -- matching the convention used when
cross-checking against SOLAR Eclipse output (see `?herit_bivar_batch`).

```{r batch, eval=FALSE}
# Illustrative only -- 'anxiety' is the only second trait in this vignette's
# toy dataset; in a real analysis this would be a panel of phenotypes.
pairs <- data.frame(
  trait1 = "sleep_quality",
  trait2 = "anxiety"
)

herit_bivar_batch(pairs, grm = A, data = dat, id_col = "IID")
```

The returned data frame has one row per pair, with `q_rhoG`, `q_rhoE`, and
`q_rhoP` alongside the raw p-values.

---

## 6. A note on validation

`herit_ace()` and `herit_bivar()` use direct numerical maximum likelihood
(Cholesky-based GLS) rather than `herit_vc()`'s eigendecomposition shortcut,
since a household term or a second trait's cross-covariance don't in general
share eigenvectors with the additive genetic matrix. Both have been validated
against real SOLAR Eclipse output on a twin dataset with a documented
ground-truth comparison -- see `NEWS.md` for the numbers and methodology, and
`?herit_bivar` for what "validated" means precisely (not bit-identical to
SOLAR, since that's not achievable between independent numerical optimisers,
but matching to several decimal places on the large majority of trait pairs
checked).

---

## 7. Citation

If you use **R-itable** in a publication, please cite:

```
Franca, L. & Leocadio-Miguel, M. (2026). R-itable: Pedigree-Based Heritability
Estimation for Family Cohort Studies. R package version 0.2.1.
https://github.com/circadia-bio/R-itable
```
